---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

# Introduction {#Introduction}

```{r Library and datasets}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,
    RColorBrewer) 
  
  theme_set(theme_cowplot())

# Scripts
# source("DTW_Script.R")
source("DTW_Script-update.R")

  
# Files  


D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern"))

  
MgCa <- read_csv(here("data","MgCa.csv"))

Ferguson_data <- read_csv(here("data","Ferguson-data.csv")) %>% 
  na.omit()

# # Check overlaps
# # Find common sample IDs
# intersect(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in MgCa
# setdiff(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in D18O
# setdiff(D18O$sampleID, MgCa$sampleID)

```

Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source. They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

Problematically, links between Mg/Ca ratios and SST changes in limpets have also been shown to be unreliable [@Graniero2015-zv] and even where they were successful, anomalous patterns in some specimens had to be filtered out, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

While recent research particularly of *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. There remains unclarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].

In this study we will repeat and expand the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which helps us to navigate the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.

Relying and adding onto published datasets, allowed us to simultaneously avoid costs for new high-resolution δ^18^O-data, use real-world examples, and also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliabilty of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.

# Materials and Methods {#Methods}

## Limpet specimens

### Modern specimens

The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.

::: {#Table_1}
| Context        | Study              | Species          | Location                                        | Sample ID     | Previous analyses                                           |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR)          | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                 | ND-1016-4     |                                                             |
|                |                    | *N. magellanica* |                                                 | NM-1016-1     |                                                             |
|                |                    |                  |                                                 | NM-1016-3     |                                                             |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | ORK-LT5       | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | QG1-7188-1    | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                 | QG1-7189-2    |                                                             |
|                |                    |                  |                                                 | QG2-1061-1    |                                                             |
|                |                    |                  |                                                 | QG2-1064-1    |                                                             |
|                |                    |                  |                                                 | QG2-7180-1    |                                                             |
|                |                    |                  |                                                 | QG2-7180-2    |                                                             |
|                |                    |                  |                                                 | QG1-7246-1    |                                                             |
|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-3 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-4 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-5 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-6 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-7 |                                                             |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-8 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-9 |                                                             |

: Overview of the modern and archaeological limpet specimens analysed in this study.
:::

```{r SST Data}

# Read the CSV files, skipping the comments
Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")

# Combine the datasets with new names and add year and month columns
SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
  mutate(analysed_sst = analysed_sst - 273.15,
         time = as_datetime(time),
         year = year(time),
         month = month(time, label = TRUE)) %>%
  group_by(dataset, year, month) %>%
  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
            sd_sst = sd(analysed_sst, na.rm = TRUE))

WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sst) %>% min() %>% round(1)
WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sst) %>% max() %>% round(1)

CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sst) %>% min() %>% round(1)
CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sst) %>% max() %>% round(1)

```

There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK). Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).

```{r SST plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from [@europeanunion-copernicusmarineservice2015]
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs


# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "Monthly Sea Surface Temperature",
       x = "Month",
       y = "Mean Analysed SST (°C)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
  labs(colour="",fill="")


```

### Archaeological specimens

In total, we studied 16 archaeological specimens. Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging. That said, the majority are under 5 years old. These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].

Nine additional shells have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.

## Methods {#methods}

### Oxygen isotopes

High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaureata* and *N. magellanica* [@Nicastro2020-ih], where detailed descriptions of the stable isotope analysis methodology can be found. In short, all shells were sectioned along the main growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromilling approach with resolutions of 100–200 µm and targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

### Mg/Ca ratios

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih]. These involve the ablation of the shell section using an infrared (1064 nm) laser pulse (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer. The resulting light spectrum quantifies the emission lines of magnesium () and calcium () to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa].

# Results {#Results}

### Modern Specimens

```{r Modern iso ranges}
# Ranges <- D18O %>% 
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>% 
#   mutate(range=max(d18O)-min(d18O),
#          range=round(range,2)) %>% 
#   select(period, sampleID,range) %>% unique()
# 
# Modern_Ranges <- Ranges %>% 
#   filter(period=="Modern") %>% 
#   arrange(range)

#Interestingly, the overall annual range in δ18O-values on Orkney (~r Modern_Ranges$range[3]‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp. shells (r min(Modern_Ranges$range)‰ to r max(Modern_Ranges$range)‰) despite the different SST ranges.

```

All modern shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp. shells living 2–3 years and *Patella v*. over 5 years. The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected barring one extreme, anomalous value in the second summer of recorded growth.

```{r Modern Isotope plot}
#| fig-cap: Isotope values of modern limpet shells from Cambaceres Bay (*a-f*) and Westray (*g*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Mod_iso
# Filter data for Modern period
modern_data <- D18O %>% filter(period == "Modern")

# Get unique sample IDs
sample_ids <- unique(modern_data$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- modern_data %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    xlim(0,15)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))
  
  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')

```

### Archaeological Specimens

```{r Archaeological Isotope plot}
#| fig-cap: Isotope values of Archaeological limpet shells from Quoygrew
#| fig-height: 10
#| fig-width: 10
#| label: fig-Arch_iso
# Filter data for Modern period
arch_data <- D18O %>% filter(period == "Archaeological")

# Get unique sample IDs
sample_ids <- unique(arch_data$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- arch_data %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))
  
  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')

```

## Patella vulgata

## Nacella deaureata

## Nacella magellanica

```{r DTW Analysis}

Samples <- unique(MgCa$sampleID)

# Initialize an empty list to store the results of shell_dtw for each sample
dtw_results <- list()

Corr_data <- tibble()

# sample <- Samples[12]

for (sample in Samples) {
  temp_D18O <- D18O %>% 
    filter(sampleID == sample) %>% 
    select(d18O) %>% 
    rename(d18o = d18O) %>% 
    mutate(sample_seq = row_number()) 

  temp_MgCa <- MgCa %>% 
    filter(sampleID == sample) %>% 
    filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>% 

    select(mg_ca) %>% 
    mutate(sample_seq = row_number())

  # Apply shell_dtw to the temp dataframes and store the result in the list
  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)
  
   # Add the sampleID column to temp_data
  temp_data <- dtw_results[[sample]]$data %>% 
    mutate(sampleID = sample)
  
  
  # Append temp_data to Corr_data
  Corr_data <- bind_rows(Corr_data, temp_data)
}


```

# 

```{r Correlation Graphs }
#| fig-cap: Correlation graphs for all specimens
#| fig-height: 20
#| fig-width: 10


P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  mutate(region = substr(sampleID, 1, 1)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
  # facet_wrap(~region,ncol=2,scales = "free")


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")


P_individual <- P_all+ylim(0,0.8)+
  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
    formula = y~x
  )+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )


P_all/P_individual+plot_layout(heights = c(0.5,0.5))


```

# Discussion

## Other correlations

```{r Ferguson Data}
#| fig-cap: Correlation graphs for Ferguson et al. specimens
#| fig-height: 20
#| fig-width: 10

Ferg_all <-
  Ferguson_data %>% 
  ggplot()+
  aes(d18o,mg_ca,col=shell)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")


Ferg_individual <- Ferg_all+
  facet_wrap(~shell,ncol=2)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 1, label.y = 26,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
    formula = y~x
  )+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )


Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))


```

| Species       | Locality       | Specimen | Correlation R^2^        | Study                             |
|---------------|---------------|---------------|---------------|---------------|
| *P. depressa* | Northern Spain | LAN541   | 0.87                    |                                   |
|               |                | LAN545   | 0.86                    |                                   |
|               |                | LAN554   | 0.78                    |                                   |
|               |                | LAN559   | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1   | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2   | 0.84                    |                                   |
|               | Crete          | AF1911A  | 0.91[^1]                |                                   |
|               |                | AF3003A  | 0.92[^2]                |                                   |
|               | Israel         | AKKPC2   | 0.96                    |                                   |
|               |                | AKKPC3   | 0.89                    |                                   |
|               |                | FRMPC1   | 0.84                    |                                   |
|               |                | FRMPC2   | 0.96                    |                                   |
|               | Libya          | MO31A    | 0.83                    |                                   |
|               |                | MP64A    | 0.33                    |                                   |
|               |                | MP67A    | 0.96                    |                                   |
|               |                | MP68A    | 0.81                    |                                   |
|               | Malta          | MA10     | 0.82                    |                                   |
|               | Tunisia        | TUNPC1   | 0.81                    |                                   |
|               |                | TUNPC2   | 0.78                    |                                   |
|               | Turkey         | ANTPC1   | 0.95                    |                                   |
|               |                | ANTPC2   | 0.93                    |                                   |
|               |                | KIZPC1   | 0.94                    |                                   |
|               |                | KIZPC2   | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
|               |                | JL2      | 0.8 (0.79)              |                                   |
| *P. caerulea* | Gibraltar      | JM00     | 0.69 (0.79)             |                                   |
|               |                | JM30     | 0.83 (0.79)             |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
|               |                |          |                         |                                   |
|               |                |          |                         |                                   |
|               |                |          |                         |                                   |

: Table 2: Overview of comparative correlations {#tab:correlations}

[^1]: SST only, no other geochemical data available

[^2]: SST only, no geochemical data available

## Comparison of ORK-LT5

```{r}

P1 <-
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  # filter(std<0.1) %>% 
  # filter(mg_ca<0.7) %>% 
  # filter(dist<20) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="firebrick")+
  # xlim(0,10)+
  facet_wrap(~sampleID,scales = "free",nrow=1)

P2 <-
  D18O %>% 
  filter(sampleID=="ORK_LT5") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path(col="cornflowerblue")+
  scale_y_reverse()+
  facet_wrap(~sampleID,scales = "free",nrow=1)

plot_compare <- P1/P2

plot_compare

```

```{r subsample}


# Function to create the desired subsampled tibble
subsample_data <- function(data, window_size = 0.05, step_size = 0.15) {
  max_dist <- max(data$dist)
  subsample_intervals <- seq(0, max_dist, by = step_size)
  
  subsampled_data <- subsample_intervals %>%
    purrr::map_df(~ {
      window_start <- .x
      window_end <- .x + window_size
      
      data_in_window <- data %>%
        filter(dist >= window_start & dist < window_end)
      
      if (nrow(data_in_window) > 0) {
        summarised_data <- data_in_window %>%
          summarise(across(c(x, y, z, mg_ca, std, relstd, dist), mean, .names = "mean_{.col}"))
        summarised_data <- summarised_data %>%
          mutate(window_start = window_start, window_end = window_end)
        return(summarised_data)
      } else {
        return(tibble())
      }
    })
  
  return(subsampled_data)
}

# Apply the function to your data
MgCa_filtered <- MgCa %>%
  filter(sampleID == "ORK_LT5")

MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)

MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)

P3.1 <-
  MgCa_subsampled_150 %>% 
ggplot()+
  aes(mean_dist,mean_mg_ca)+
  geom_path(col="firebrick2")

P3.2 <-
  MgCa_subsampled_300 %>% 
ggplot()+
  aes(mean_dist,mean_mg_ca)+
  geom_path(col="firebrick4")

P1/P3.1/P3.2

```

```{r}
# This is data extracted from Fig. 6E of Graniero et al. 2016
GranieroDF <- tibble::tribble(
                ~dist, ~mg_ca,
                  0.1,     19,
                  0.2,   20.6,
                  0.4,   23.8,
                  0.6,   20.4,
                    1,   19.1,
                  1.3,   14.9,
                  1.6,     16,
                  1.9,   17.8,
                  2.3,   15.8,
                  2.8,   17.3,
                  3.1,   13.8,
                  3.3,   24.9,
                  3.7,   16.1,
                    4,   15.6,
                  4.3,   18.4,
                  4.6,   15.5,
                  5.2,   18.6,
                  5.2,   21.2,
                  5.5,   24.1,
                  5.8,   18.7,
                  6.2,   19.3,
                  6.3,   15.5,
                  6.7,   16.7,
                  6.9,   15.3,
                  7.3,   17.9,
                  7.6,   13.7,
                    8,   11.8,
                  8.2,   16.2,
                  8.5,   18.1,
                  8.9,     18,
                  9.1,   19.7,
                  9.4,   20.8,
                  9.7,   19.6,
                   10,   22.2,
                 10.2,   20.2,
                 10.5,   19.3,
                 10.8,   22.8,
                 11.2,   18.7,
                 11.8,   22.1
                )

```

<!--# Maybe get the actual data from the image itself. -->
