---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata*, and *N. magellanica* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

```{r Library and datasets}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#knitr::opts_chunk$set(fig.path = "")
knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,
    RColorBrewer) 
  
  theme_set(theme_cowplot(10))

# Scripts
# source("DTW_Script.R")
source("DTW_Script-update.R")

  
# Files  


D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern")) %>% 
  mutate(species = case_when(
      startsWith(sampleID, "ND") ~ "Nacella deaurata",
      startsWith(sampleID, "NM") ~ "Nacella magellanica",
      TRUE ~ "Patella vulgata"
    ))

  
MgCa <- read_csv(here("data","MgCa.csv"))

Ferguson_data <- read_csv(here("data","Ferguson-data.csv")) %>% 
  na.omit()

# # Check overlaps
# # Find common sample IDs
# intersect(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in MgCa
# setdiff(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in D18O
# setdiff(D18O$sampleID, MgCa$sampleID)

D18O <-   D18O %>% filter(sampleID %in% MgCa$sampleID)

```

# Introduction {#Introduction}

Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz]. They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2017-io; @Poulain2015-dg; @Vihtakari2017-wd]. This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as a temperature proxy. Confusingly, in some cases multiple different temperature equations have been found for the same species (see Freitas *et al*. [-@Freitas2012-tx] and Vihtakari *et al*. [-@Vihtakari2017-wd] and references therein). Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi]. Where the use of Mg/Ca ratios as a palaeotemperature proxy was successful, anomalous patterns in some specimens had to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

Recent research particularly of *Patella* spp. in the Mediterranean and southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io].

In this study we will repeat and expand the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which allows navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio. By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.

# Materials and Methods {#Methods}

## Limpet specimens

The analysed specimens, their origin and respective studies with research background information can be found in Table [1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not allow them to be re-analysed.

```{r SST_Data}

# Read the CSV files, skipping the comments
Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")

# Combine the datasets with new names and add year and month columns
SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
  mutate(analysed_sst = analysed_sst - 273.15,
         time = as_datetime(time),
         year = year(time),
         month = month(time, label = TRUE)) %>%
  group_by(dataset, year, month) %>%
  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
            sd_sst = sd(analysed_sst, na.rm = TRUE))

WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sst) %>% min() %>% round(1)
WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sst) %>% max() %>% round(1)

CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sst) %>% min() %>% round(1)
CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sst) %>% max() %>% round(1)

```

::: {#Table_1}
\fontsize{8pt}{8pt}\selectfont

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|------|------------|-------|--------|------|----------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C        |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C       |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C       |
|                |                    |                  | Westray (UK)    | QG1-7189-2 |       analysis                        |
|                |                    |                  |                 | QG2-1061-1 |                                       |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |

: Overview of the modern and archaeological limpet specimens analysed in this study.
:::

\normalsize

Specimens were sourced from two main locations: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in the Orkney archipelago of Scotland (UK). Four modern specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm of precipitation, but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units) [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs -- red line).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$^13^C values, as well as LA-ICP-MS [@Graniero2017-io]. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).

```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl.
#| fig-height: 3
#| fig-width: 6
#| label: fig-SSTs



# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "",
       x = "Month",
       y = "Mean Analysed SST (°C)",colour="",fill="") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) 

```

We also studied 7 archaeological specimens of *P. vulgata* from the Quoygrew site on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates. That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.

## Methods {#methods}

### Mg/Ca ratios

Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih]. The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs). The resulting light spectrum quantifies the emission lines of Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate. All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.

Using this system we carried out elemental imaging of the shell specimens at a resolution of 50--100 µm distance between sampling spots. Each spot was irradiated 10 times with the first 3 spectra discarded as a cleaning step and the remaining spectra summed to get an average Mg/Ca intensity ratio for each sample spot. Subsequently, we re-sampled the section using a line scan at 10 µm resolution. This leads to an overlap between sample spots, but also allows for a continuous record without gaps. Intensity ratios were filtered for cases with high relative standard deviation (i.e., more than 10%). These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surface introducing variability in the plasma generation and thus uncertainty in the data of these locations.

### Oxygen isotope ratios

High-quality $\delta$^18^O values were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaurata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found. In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

```{r Nacella_Isotope_plot,dev='png',dpi=300}
#| fig-cap: Isotope values of *Nacella* spp. limpet shells from Cambaceres Bay.
#| fig-height: 5
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

# Get unique sample IDs
sample_ids <- unique(Nacella_iso$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(paste(delta^{18},"O (\u2030)")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))
  
  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=2) + plot_annotation(tag_levels = 'a')

```

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively). Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp. shells living 2–3 years and *P. vulgata* living regularly over 3 years. In those instances where longer lifespans resulted in slow growth rates below 1 mm/yr (i.e. specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7.

```{r Patella_Isotope_plot,dev='png',dpi=300,fig.pos='H'}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

# Get unique sample IDs
sample_ids <- unique(Patella_iso$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Patella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))
  
  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')

```

## Dynamic Time Warping

Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_Surge_Godino_2024]). The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.

```{r DTW Analysis}

Samples <- unique(MgCa$sampleID)

# Initialize an empty list to store the results of shell_dtw for each sample
dtw_results <- list()

Corr_data <- tibble()

# sample <- Samples[12]

for (sample in Samples) {
  temp_D18O <- D18O %>% 
    filter(sampleID == sample) %>% 
    select(d18O) %>% 
    rename(d18o = d18O) %>% 
    mutate(sample_seq = row_number()) 

  temp_MgCa <- MgCa %>% 
    filter(sampleID == sample) %>% 
    filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>% 

    select(mg_ca) %>% 
    mutate(sample_seq = row_number())

  # Apply shell_dtw to the temp dataframes and store the result in the list
  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)
  
   # Add the sampleID column to temp_data
  temp_data <- dtw_results[[sample]]$data %>% 
    mutate(sampleID = sample)
  
  
  # Append temp_data to Corr_data
  Corr_data <- bind_rows(Corr_data, temp_data)
}


```

# Results {#Results}

## *Patella vulgata*

The elemental imaging of *P. vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating high temperature periods and low ratios indicating low temperature periods. As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries). The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2)(see Fenger et al. [-@Fenger2007-gf] for a detailed explanations of microstructural layers in *P. vulgata*). The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above). Compared to other patelloid shells, some of whose interior is almost entirely made of aragonite, this layer is very thin. \newline

```{r Patella_LIBS,fig.pos='H'}
#| fig-cap: Example LIBS maps and line scans of analysed specimens. (*a+c*) ORK-LT and (*b*+*d*) QG1-7188-1.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
#| fig-height: 6
#| fig-width: 6
#| label: fig-Pat_LIBS

#ORK_LT5
ORK_LT5_map <-
  read_csv(here("data", "map_data", "ORK_LT_5_A100_All_analyzed_data.csv"))

ORK_LT5_plot <- 
  ORK_LT5_map %>%
  filter(std < 0.15) %>%
 #filter(x>0) %>% 
  filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(data = ORK_LT5_map, fill = "grey") +
  geom_tile(show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="ORK_LT5"),aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(8)+
  scale_fill_viridis_c(limits = c(0.3, 0.7),
                       breaks = seq(0.3, 0.7, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", fill = "Mg/Ca ", subtitle="ORK-LT5")
    
ORK_LT5_line <- 
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(8)+
    theme(plot.tag.position = c(-0.01, 1))+ # Adjust horizontal and vertical position
  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
  
  
  
###########   
#QG1_7188_1
###########
QG1_7188_1_map <-
  read_csv(here("data", "map_data", "QG1-7188-1-AB50_analyzed_data.txt"))

QG1_7188_1_plot <- 
  QG1_7188_1_map %>%
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, col = mg_ca) +
  geom_point(data = QG1_7188_1_map, col = "grey",size=0.2) +
  geom_point(size=0.2,show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="QG1-7188-1"),aes(x,y),col="white",linetype="longdash")+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(8)+
  theme(
    legend.position = c(0.05, 0.78),
   legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", colour = "Mg/Ca", subtitle="QG1-7188-1")

QG1_7188_1_line <- 
  MgCa %>% 
  filter(sampleID=="QG1-7188-1") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(8)+
  theme(plot.tag.position = c(-0.01, 1))+ # Adjust horizontal and vertical position
  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
  
  
  
################
####FIGURE######
################
#(ORK_LT5_plot+ORK_LT5_line)/(QG1_7188_1_plot+QG1_7188_1_line)+plot_annotation(tag_levels = "a")

#QG1_7188_1_line/ORK_LT5_line/
maps <-     (ORK_LT5_plot+labs(tag="a"))+(QG1_7188_1_plot+labs(tag = "b"))+plot_layout(widths = c(0.4,0.2))#heights=c(0.55,1.2))

lines <-   (ORK_LT5_line+labs(tag="c"))/(QG1_7188_1_line+labs(tag = "d"))+plot_layout(heights=c(0.5,0.5))

maps/lines+plot_layout(heights=c(0.6,0.5))

```

Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability. This is particularly visible in @fig-Pat_LIBS b, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis). A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).

Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental imaging. That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D image. This shows the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

```{r Patella_Compare,dev='png',dpi=300}
#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of specimen ORK-LT5 
#| fig-height: 7
#| fig-width: 16
#| label: fig-Pat_Comp
 
P1 <-
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(17)+
  labs(x = "", y = "Mg/Ca intensity ratio",subtitle="ORK-LT5 ")



P2 <-
  D18O %>% 
  filter(sampleID=="ORK_LT5") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(18)+
  scale_y_reverse()+
  labs(x = "Distance to edge (mm)", y = expression(paste(delta^{18},"O (\u2030)")))


P1/P2+plot_annotation(tag_levels = "a")

```

Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value of the individual *P. vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.

```{r Patella_Correlation_Graphs,dev='png',dpi=300}
#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
#| fig-height: 8
#| fig-width: 16
#| label: fig-Pat_Corr


P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  # mutate(region = substr(sampleID, 1, 1)) %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(species=="Patella vulgata") %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  theme_cowplot(20)+
  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c("#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063")) +
scale_fill_manual(values = c( "#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063"))#+
  # facet_wrap(~region,ncol=2,scales = "free")


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")

P_individual <- P_all+ylim(0.2,0.8)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )



#P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
P_individual

```

## *Nacella* spp.

The LIBS data for *Nacella* spp. specimens is less straightforward with shells being much thinner than *P. vulgata* and thus patterns being more difficult to discern. Both specimens (ND-1016-3 and NM-1016-3) shown as examples in @fig-Nac_LIBS (see additional graphs in supplementaries) also experience increases of Mg/Ca intensity ratio towards the exterior of the shell, similar to the *P. vulgata* specimens from Orkney. This intra-increment heterogeneity combined with the thinness of the shell, further complicates their analysis compared to *P. vulgata* or other *Patella* species [@Hausmann2019-fi]. Nevertheless, repeating patterns were visible in the anterior section of the shell, which were also mirrored in the inner layers (m-2) of specimens NM-1016-3 (@fig-Nac_LIBS b).

```{r Nacella_LIBS,fig.pos='H'}
#| fig-cap: Example LIBS maps and line scans of analysed *Nacella* spp. specimens. (*a+c*) ND-1016-3 and (*b*+*d*) NM-1016-3.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range. Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
#| fig-height: 7
#| fig-width: 16
#| label: fig-Nac_LIBS

#ORK_LT5
ND_1016_map <-
  read_csv(here("data", "map_data", "ND-106-3-A100B_analyzed_data.csv"))

ND_Line <- read_csv(here("data", "map_data", "ND-106-3-A10-LineScan_analyzed_data.csv")) %>% 
  mutate(dist=row_number()*0.01)


ND_1016_plot <- 
  ND_1016_map %>%
 # filter(std < 0.15) %>%
  #filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(data = ND_1016_map, fill = "grey") +
  geom_tile(show.legend=FALSE) +
  geom_path(data=ND_Line,aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0.05, 0.45),
                       breaks = seq(0, 0.5, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", fill = "Mg/Ca ",subtitle="ND-1016-3")
    
ND_1016_line <- 
  ND_Line %>% 
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c() +    
  theme_cowplot(15)+
  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio",subtitle="ND-1016-3")
  
  
  
###########
#NM-1016-3
###########
NM_1016_map100 <-
  read_csv(here("data", "map_data", "NM-1016-3-A100_analyzed_data.csv"))
NM_1016_map50 <-
  read_csv(here("data", "map_data", "NM-1016-3-A50Edge_analyzed_data.csv"))

NM_Line <- read_csv(here("data", "map_data", "NM-1016-3-A10LineScan2_analyzed_data.csv")) %>% 
  mutate(dist=row_number()*0.01)

NM_1016_plot100 <- 
  NM_1016_map100 %>%
  filter(std < 0.15) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(show.legend=FALSE,size=0.5) +
  geom_path(data=NM_Line,aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0.1, 0.4),
                       breaks = seq(0, 0.4, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
   annotate("rect", 
           xmin = min(NM_1016_map50$x)+15, 
           xmax = max(NM_1016_map50$x)+15, 
           ymin = min(NM_1016_map50$y)-2.3, 
           ymax = max(NM_1016_map50$y)-2, 
           alpha = 0, linetype="longdash",col = "black")+
    labs(x = "mm", y = "mm", fill = "Mg/Ca ",subtitle="NM-1016-3")

NM_1016_plot50 <- 
  NM_1016_map50 %>%
 # filter(std < 0.15) %>%
  #filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(show.legend=FALSE,size=0.5) +
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0.05, 0.4),
                       breaks = seq(0, 0.4, by = 0.1)) +    
theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )+ 
    coord_fixed() +
    labs(x = "mm", y = "mm", colour = "Mg/Ca ")
    
NM_1016_line <- 
  NM_Line %>% 
  filter(relstd<0.1) %>% 
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c() +    
  theme_cowplot(15)+
  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio",subtitle="NM-1016-3")
  
  
NM_1016_plot <- NM_1016_plot100+inset_element(NM_1016_plot50+theme(plot.tag = element_blank()), 0, 0, 0.7, 0.7)
  
  
################
####FIGURE######
################
Maps <- ND_1016_plot|NM_1016_plot
Lines <- ND_1016_line/NM_1016_line
Maps+Lines+ plot_annotation(tag_levels = list(c("a", "b", "","c","d")))


#(ND_1016_plot+ND_1016_line)/(NM_1016_plot+NM_1016_line)+plot_annotation(tag_levels = "a")

```

Interestingly, the Mg/Ca intensity ratios in the *Nacella* spp. shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g. 0.3–0.7 in *P. vulgata* above, or 0.5–1.5 in Hausmann *et al*. [-@Hausmann2023-ih]) using the same emission lines. While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the chosen Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere or seen only in aragonitic parts of e.g. *Patella* shells. Similar observations have been made before [@Graniero2017-io]. Strong intra-increment heterogeneities are visible along the exterior of the shells (m+3 layer), where Mg/Ca intensity ratios are higher, particularly in specimen ND-1016-3 (@fig-Nac_LIBS). The step from lower to higher along the increment is not gradual, as is the case in NM-1016-3 (see inset of @fig-Nac_LIBS b), but rather stepwise. Sampling across this boundary in a linear scan would most likely lead to erroneous results. Also sampling across gradual changes as in NM-1016-3 can influence the overall correlation graph for a specimen.

Line scans reflect well the changes previously indicated by $\delta$^18^O values and also indicate about two years of recorded growth. That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.

@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O values (similar graphs for the other *Nacella* spp. specimens can be found in the supplementaries). Compared to the *P. vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and thus a lower chance of misaligning annual extremes.

```{r Nacella_Compare,dev='png',dpi=300}
#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of *Nacella* spp. specimens. 
#| fig-height: 7
#| fig-width: 16
#| label: fig-Nac_Comp
 
P1 <-
  ND_Line %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6) %>%
  filter(dist<10) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(15)+
  labs(x = "", y = "Mg/Ca intensity ratio", subtitle="ND-1016-3")



P2 <-
  D18O %>% 
  filter(sampleID=="ND-1016-3") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(15)+
  scale_y_reverse()+
  labs(x = "Distance to edge (mm)", y = expression(paste(delta^{18},"O (\u2030)")))


P3 <-
  NM_Line %>% 
  filter(relstd < 0.1) %>%
  filter(dist<9.5) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(15)+
  labs(x = "", y = "Mg/Ca intensity ratio", subtitle="NM-1016-3")



P4 <-
  D18O %>% 
  filter(sampleID=="NM-1016-3") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(15)+
  scale_y_reverse()+
  labs(x = "Distance to edge (mm)", y = expression(paste(delta^{18},"O (\u2030)")))


 ((P1 / P2) | (P3 / P4)) + plot_annotation(tag_levels = 'a')

```

The *Nacella* spp. specimens showed similarly clustered correlations, which interestingly were not grouped by species. In fact, specimens ND-1016-3 (*Nacella deaurata)* and NM-1016-3 (*N. magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr). Both also share a high coefficient of determination (R^2^) of 0.95 and 0.92 respectively, with the remaining specimens being somewhat lower (R^2^ = 0.76 for ND-1016-4 and R^2^=0.84 for NM-1016-3). The mean R^2^ value for each *N.* species is 0.86 for *N. deaurata* and 0.88 for *N. magellanica*. Why two specimens of different species are so well aligned is not entirely certain and randomness cannot for now be ruled out. That said, since they both have a high R^2^ value, it might just be that they were both minimally affected by factors other than SST, including the sampling location of our line scans.

```{r Nacella_Correlation_Graphs,dev='png',dpi=300}
#| fig-cap: Correlation graphs for *Nacella* spp. specimens
#| fig-height: 5
#| fig-width: 16
#| label: fig-Nac_Corr


P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(grepl("Nacella", species)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  theme_cowplot(10)+
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
 # ylim(0.05,NA)+
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12"))#+
  # facet_wrap(~region,ncol=2,scales = "free")


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")

Nacella_Corr_data <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(grepl("Nacella", species))

P_individual <- P_all+ylim(0,0.4)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
    coord_fixed(ratio = 6) +
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

P_individual
#P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")


```

# Discussion {#Discussion}

The Mg/Ca ratios from the 2D elemental imaging present positive results suggesting that SST is the main factor influencing variations in Mg/Ca ratios in three Atlantic limpet species. Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data. Nevertheless, the resulting R^2^-values for each species were high (*P. vulgata mean R^2^*=0.89, *N. deaurata mean R^2^*=0.86*, N. magellanica mean R^2^*=0.88). These results are similar to previously published data on other *Patella* species shells (Table [2](#Table_2)): R^2^ values of *P. depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P. caerulea* from the Mediterranean between 0.78 and 0.96 — barring one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].

::: {#Table_2}
\fontsize{8pt}{8pt}\selectfont

| Species       | Locality       | Specimen  | R^2^                    | Study                             |
|------|------------|-------|--------|------|----------|
| *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
|               |                | LAN545    | 0.86                    |                                   |
|               |                | LAN554    | 0.78                    |                                   |
|               |                | LAN559    | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1    | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2    | 0.84                    |                                   |
|               | Crete          | AF1911A   | 0.91                    |                                   |
|               |                | AF3003A   | 0.92                    |                                   |
|               | Israel         | AKKPC2    | 0.96                    |                                   |
|               |                | AKKPC3    | 0.89                    |                                   |
|               |                | FRMPC1    | 0.84                    |                                   |
|               |                | FRMPC2    | 0.96                    |                                   |
|               | Libya          | MO31A     | 0.83                    |                                   |
|               |                | **MP64A** | **0.33**                |                                   |
|               |                | MP67A     | 0.96                    |                                   |
|               |                | MP68A     | 0.81                    |                                   |
|               | Malta          | MA10      | 0.82                    |                                   |
|               | Tunisia        | TUNPC1    | 0.81                    |                                   |
|               |                | TUNPC2    | 0.78                    |                                   |
|               | Turkey         | ANTPC1    | 0.95                    |                                   |
|               |                | ANTPC2    | 0.93                    |                                   |
|               |                | KIZPC1    | 0.94                    |                                   |
|               |                | KIZPC2    | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |
|               |                | JL2       | 0.8\*\*                 |                                   |
| *P. caerulea* | Gibraltar      | JM00      | 0.69\*\*                |                                   |
|               |                | JM30      | 0.83\*\*                |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5   | not reported, here 0.88 | [@Graniero2017-io] and this study |

: Correlations of Mg/Ca ratios with $\delta$^18^O values and SST found in other studies. Bold R^2^ values are interpreted as outliers. \* indicates that only SST data and no other geochemical data was used. \*\* these samples were used to determine a shared R^2^-value of 0.79.
:::

\normalsize

*P. rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.80, respectively [@Ferguson2011-zl] (@fig-Ferg). Together with the data from two *P. caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used to determine a general Mg/Ca-SST relationship with an R^2^ value of 0.79. JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data. While no additional information about those specimens is available, it seems likely that the discarded Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well. Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could have potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset. Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.

```{r Ferguson Data, dev='png'}
#| fig-cap: Correlation graphs for Ferguson et al., (2011) specimens.
#| fig-height: 5
#| fig-width: 16
#| label: fig-Ferg

Ferg_all <-
  Ferguson_data %>% 
  ggplot()+
  aes(d18o,mg_ca,col=shell)+
  geom_point(show.legend = FALSE)+
  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca (mmol/mol)")+
  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))

Ferg_individual <- Ferg_all+
  facet_wrap(~shell,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.3, label.y = 26,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+  
  theme_cowplot(20)+
  coord_fixed(ratio =0.22) +
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )
Ferg_individual
#Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
```

Another shell whose results we aim to explain further is ORK-LT5 (@fig-Pat_LIBS), which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io]. In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub). These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August). Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88 (@fig-ORK_sub).

```{r ORK_LT5}
#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and $\delta$^18^O-values
#| fig-height: 4.5
#| fig-width: 15
#| label: fig-ORK_sub


#
# This is data extracted from Fig. 6E of Graniero et al. 2016
GranieroDF <- tibble::tribble(
                ~dist, ~mg_ca,
                  0.1,     19,
                  0.2,   20.6,
                  0.4,   23.8,
                  0.6,   20.4,
                    1,   19.1,
                  1.3,   14.9,
                  1.6,     16,
                  1.9,   17.8,
                  2.3,   15.8,
                  2.8,   17.3,
                  3.1,   13.8,
                  3.3,   24.9,
                  3.7,   16.1,
                    4,   15.6,
                  4.3,   18.4,
                  4.6,   15.5,
                  5.2,   18.6,
                  5.2,   21.2,
                  5.5,   24.1,
                  5.8,   18.7,
                  6.2,   19.3,
                  6.3,   15.5,
                  6.7,   16.7,
                  6.9,   15.3,
                  7.3,   17.9,
                  7.6,   13.7,
                    8,   11.8,
                  8.2,   16.2,
                  8.5,   18.1,
                  8.9,     18,
                  9.1,   19.7,
                  9.4,   20.8,
                  9.7,   19.6,
                   10,   22.2,
                 10.2,   20.2,
                 10.5,   19.3,
                 10.8,   22.8,
                 11.2,   18.7,
                 11.8,   22.1
                )

P3.1 <-  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca)+
  geom_path(show.legend = FALSE)+
    scale_x_continuous(breaks = seq(0,12,by=2),
                     labels = seq(0,12,by=2))+
  theme_cowplot(12)+
    annotate("rect", xmin = 1.5, xmax = 4, ymin = 0.3, ymax = 0.55, alpha = 0.2, fill = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.3, ymax = 0.55, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
  

P3.3 <-
  GranieroDF %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="black")+
   theme_cowplot(12)+
  scale_x_continuous(breaks = seq(0,12,by=2),
                     labels = seq(0,12,by=2))+
  annotate("rect", xmin = 1.5, xmax = 3, ymin = 10, ymax = 25, alpha = 0.2, fill = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.08, ymin = 10, ymax = 25, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
  labs(x = "Distance to edge (mm)", y = "Mg/Ca (mmol/mol)")
  

P3.1/P3.3

```

We were not able to resample the *Nacella* spp. specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of $\delta$^18^O values and Mg/Ca ratios and pointed to no relationship between elemental ratios and SST at all. However, we were able to source other modern specimens of the same species from Tierra del Fuego, which shall serve here as an indicator. Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^ values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries). Why these results are different from the studied *Nacella* spp. specimens in Graniero *et al*. [-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in Graniero *et al*. [-@Graniero2017-io]) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — is not perfectly visible, leading to an overall result of an uncorrelated relationship between Mg/Ca ratios and SST.

# Conclusion

This study uses LIBS as a novel analytical tool to present high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relationship to sea surface temperature. The results for *P. vulgata*, *N. deaurata*, and *N. magellanica* were consistently promising with mean R^2^ values of 0.89, 0.86, and 0.88 respectively.

LIBS elemental imaging provided insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the $\delta$^18^O values of the same shells. Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.

These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past. Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl]. By using LIBS as a screening tool for targeted $\delta$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency. This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.


# Acknowledgements


# References

```{r}
#header stuff
#include-in-header:
#  text: |
#    \usepackage{graphicx}
#    \usepackage{booktabs}
#    \usepackage{amsmath}
#    \usepackage{unicode-math}
#    \setmathfont{XITS Math}
```
