---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

```{r Library and datasets}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,kableExtra,
    RColorBrewer) 
  
  theme_set(theme_cowplot())

# Scripts
# source("DTW_Script.R")
source("DTW_Script-update.R")

  
# Files  


D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern")) %>% 
  mutate(species = case_when(
      startsWith(sampleID, "ND") ~ "Nacella deaureata",
      startsWith(sampleID, "NM") ~ "Nacella magellanica",
      TRUE ~ "Patella vulgata"
    ))

  
MgCa <- read_csv(here("data","MgCa.csv"))

Ferguson_data <- read_csv(here("data","Ferguson-data.csv")) %>% 
  na.omit()

# # Check overlaps
# # Find common sample IDs
# intersect(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in MgCa
# setdiff(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in D18O
# setdiff(D18O$sampleID, MgCa$sampleID)

D18O <-   D18O %>% filter(sampleID %in% MgCa$sampleID)

```

# Introduction {#Introduction}

Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source. They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2015-zv; @Poulain2015-dg; @Vihtakari2017-wd]. This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy. Confusingly, multiple different temperature equations have been found for the same species (see [@Freitas2012-tx; @Vihtakari2017-wd] and references therein). Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi]. Where the use of Mg/Ca as a palaeotemperature proxy was successful, anomalous patterns in some specimens had still to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

While recent research particularly of *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. There remains a lack of clarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].

In this study we will repeat and expand the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies. To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS. LIBS allows us to carry out 2D imaging of entire shell sections, which helps us to navigate the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.

By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution δ^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.

# Materials and Methods {#Methods}

## Limpet specimens

### Modern specimens

The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.

::: {#Table_1}
| Context        | Study              | Species          | Location                                        | Sample ID     | Previous analyses                                           |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR)          | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                 | ND-1016-4     |                                                             |
|                |                    | *N. magellanica* |                                                 | NM-1016-1     |                                                             |
|                |                    |                  |                                                 | NM-1016-3     |                                                             |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | ORK-LT5       | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | QG1-7188-1    | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                 | QG1-7189-2    |                                                             |
|                |                    |                  |                                                 | QG2-1061-1    |                                                             |
|                |                    |                  |                                                 | QG2-1064-1    |                                                             |
|                |                    |                  |                                                 | QG2-7180-1    |                                                             |
|                |                    |                  |                                                 | QG2-7180-2    |                                                             |
|                |                    |                  |                                                 | QG1-7246-1    |                                                             |
|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-3 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-4 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-5 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-6 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-7 |                                                             |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-8 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-9 |                                                             |

: Overview of the modern and archaeological limpet specimens analysed in this study.
:::

```{r SST_Data}

# Read the CSV files, skipping the comments
Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")

# Combine the datasets with new names and add year and month columns
SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
  mutate(analysed_sst = analysed_sst - 273.15,
         time = as_datetime(time),
         year = year(time),
         month = month(time, label = TRUE)) %>%
  group_by(dataset, year, month) %>%
  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
            sd_sst = sd(analysed_sst, na.rm = TRUE))

WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sst) %>% min() %>% round(1)
WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
  pull(mean_sst) %>% max() %>% round(1)

CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sst) %>% min() %>% round(1)
CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
  pull(mean_sst) %>% max() %>% round(1)

```

There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK). Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009. Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho]. The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).

```{r SST_plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs



# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "Monthly Sea Surface Temperature",
       x = "Month",
       y = "Mean Analysed SST (°C)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
  labs(colour="",fill="")


```

### Archaeological specimens

In total, we studied 16 archaeological specimens. Of those, seven *P. vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging. That said, the majority are under 5 years old. These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].

Nine additional shells of *N. magellanica* (n=6) and *N*. *deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells. They derive from the site of Heshkaia 35 situated within Moat Bay, around 30 km east of where the modern specimens in our study were collected [@Zangrando2014-qi; @Zangrando2021-qg]. With radiocarbon dates putting the shells into the Late Holocene (1300 – 1450 CE) they are of similar age — if somewhat younger — as the assemblage from Quoygrew.

## Methods {#methods}

### Mg/Ca ratios

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih]. These involve the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer. The resulting light spectrum quantifies the emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII; 315.887 nm) to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca changes within the shell carbonate.

Using this system we carried out elemental imaging of the shell specimens at a resolution of 50 µm distance between sampling spots. Each spot was irradiated 10 times with the 3 first spectra used for cleaning steps and the remaining summed to get an average Mg/Ca intensity ratio for each sample spot. Subsequently, we re-sampled the section using a line scan at 10 µm resolution. This leads to an overlap between sample spots, but also allows for a continuous record without gaps. Intensity ratios were filtered for cases with high relative standard deviation (i.e. more than 10%). These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surfaces introducing variability in the plasma generation and thus uncertainty in the data of these locations.

### Oxygen isotopes

High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaureata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found. In short, all shells were sectioned along the main growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species. Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp. shells living 2–3 years and *Patella v*. regularly over 3 years. In those instances where longer lifetimes are encountered (i.e. specimen QG1-7246-1, @fig-Pat_iso b). Interestingly, the annual ranges in δ^18^O-values are not similar between the coeval *Nacella* sp. specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰. *P. vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

# Get unique sample IDs
sample_ids <- unique(Nacella_iso$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))
  
  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=3) + plot_annotation(tag_levels = 'a')

```

```{r Patella_Isotope_plot}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

# Get unique sample IDs
sample_ids <- unique(Patella_iso$sampleID)

# Initialise an empty list to store plots
plots <- list()

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Patella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))
  
  # Store plot in the list
  plots[[i]] <- plot
}

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')

```

## Dynamic Time Warping

Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of δ^18^O-values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy data sets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment. The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets. The code used for this analysis as well as the data required to run it can be found in this **OSF repository**.

```{r DTW Analysis}

Samples <- unique(MgCa$sampleID)

# Initialize an empty list to store the results of shell_dtw for each sample
dtw_results <- list()

Corr_data <- tibble()

# sample <- Samples[12]

for (sample in Samples) {
  temp_D18O <- D18O %>% 
    filter(sampleID == sample) %>% 
    select(d18O) %>% 
    rename(d18o = d18O) %>% 
    mutate(sample_seq = row_number()) 

  temp_MgCa <- MgCa %>% 
    filter(sampleID == sample) %>% 
    filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>% 

    select(mg_ca) %>% 
    mutate(sample_seq = row_number())

  # Apply shell_dtw to the temp dataframes and store the result in the list
  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)
  
   # Add the sampleID column to temp_data
  temp_data <- dtw_results[[sample]]$data %>% 
    mutate(sampleID = sample)
  
  
  # Append temp_data to Corr_data
  Corr_data <- bind_rows(Corr_data, temp_data)
}


```

# Results {#Results}

## *Patella vulgata*

The elemental imaging of *P. vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating and low ratios indicating low temperature periods. As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries). The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2) [@Fenger2007-gf]. The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above). Compared to other patelloid shells, whose interior is almost entirely made of aragonite, this layer is very thin.

```{r Patella_LIBS}
#| fig-cap: Example LIBS maps and line scans of analysed specimens. (*a+b*) ORK-LT and (*c*+*d*) QG1-7188-1.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
#| fig-height: 10
#| fig-width: 15
#| label: fig-Pat_LIBS

#ORK_LT5
ORK_LT5_map <-
  read_csv(here("data", "map_data", "ORK_LT_5_A100_All_analyzed_data.csv"))

ORK_LT5_plot <- 
  ORK_LT5_map %>%
  filter(std < 0.15) %>%
  filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(data = ORK_LT5_map, fill = "grey") +
  geom_tile(show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="ORK_LT5"),aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0.3, 0.7),
                       breaks = seq(0.3, 0.7, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
    
ORK_LT5_line <- 
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(20)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
  
  
  
###########   
#QG1_7188_1
###########
QG1_7188_1_map <-
  read_csv(here("data", "map_data", "QG1-7188-1-AB50_analyzed_data.txt"))

QG1_7188_1_plot <- 
  QG1_7188_1_map %>%
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, col = mg_ca) +
  geom_point(data = QG1_7188_1_map, col = "grey") +
  geom_point(size=1,show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="QG1-7188-1"),aes(x,y),col="white",linetype="longdash")+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(20)+
  theme(
    legend.position = c(0.05, 0.78),
   legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", colour = "Mg/Ca")

QG1_7188_1_line <- 
  MgCa %>% 
  filter(sampleID=="QG1-7188-1") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(20)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
  
  
  
################
####FIGURE######
################
  (ORK_LT5_plot+ORK_LT5_line)/(QG1_7188_1_plot+QG1_7188_1_line)+plot_annotation(tag_levels = "a")
```

Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability. This is particularly visible in @fig-Pat_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis). A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).

Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging. That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (\@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its δ^18^O-values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances. Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

```{r Patella_Compare}
#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
#| fig-height: 10
#| fig-width: 20
#| label: fig-Pat_Comp
 
P1 <-
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(20)+
  labs(x = "", y = "Mg/Ca intensity ratio")



P2 <-
  D18O %>% 
  filter(sampleID=="ORK_LT5") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(20)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))


P1/P2+plot_annotation(tag_levels = "a")

```

Aligning the records programmatically using dynamic time warping, allowed us to compare them directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (δ^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest, respectively, specimens, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value is 0.89, suggesting a generally good fit between the two proxies.

```{r Patella_Correlation_Graphs }
#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
#| fig-height: 18
#| fig-width: 15
#| label: fig-Pat_Corr


P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  # mutate(region = substr(sampleID, 1, 1)) %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(species=="Patella vulgata") %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  theme_cowplot(20)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c("#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063")) +
scale_fill_manual(values = c( "#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063"))#+
  # facet_wrap(~region,ncol=2,scales = "free")


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")

P_individual <- P_all+ylim(0,0.8)+
  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )



P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")


```

## *Nacella* sp.

The LIBS data for *Nacella* sp. specimens is less straightforward with shells being much thinner and thus patterns being more difficult to make out. Both specimens (ND-1016-3 and NM-1016-3) shown as example in @fig-Nac_LIBS (see additional graphs in supplementaries) also experience increases of Mg/Ca intensity ratio towards the exterior of the shell, similar to the *P. vulgata* specimens from Orkney. This intra-increment heterogeneity combined with the thinness of the shell, further complicates their analysis compared to *P. vulgata* or other *Patella* species [@Hausmann2019-fi]. Nevertheless, repeating patterns were visible in the anterior section of the shell, which were also mirrored in the inner layers (m-2) of specimens NM-1016-3 (@fig-Nac_LIBS c).

```{r Nacella_LIBS}
#| fig-cap: Example LIBS maps and line scans of analysed specimens. (*a+b*) ND-1016-3 and (*c*+*d*) NM-1016-3.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range. Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
#| fig-height: 10
#| fig-width: 15
#| label: fig-Nac_LIBS

#ORK_LT5
ND_1016_map <-
  read_csv(here("data", "map_data", "ND-106-3-A100B_analyzed_data.csv"))

ND_Line <- read_csv(here("data", "map_data", "ND-106-3-A10-LineScan_analyzed_data.csv")) %>% 
  mutate(dist=row_number()*0.01)


ND_1016_plot <- 
  ND_1016_map %>%
 # filter(std < 0.15) %>%
  #filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(data = ND_1016_map, fill = "grey") +
  geom_tile(show.legend=FALSE) +
  geom_path(data=ND_Line,aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0, 0.5),
                       breaks = seq(0, 0.5, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
    
ND_1016_line <- 
  ND_Line %>% 
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c() +    
  theme_cowplot(20)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
  
  
  
###########
#NM-1016-3
###########
NM_1016_map100 <-
  read_csv(here("data", "map_data", "NM-1016-3-A100_analyzed_data.csv"))
NM_1016_map50 <-
  read_csv(here("data", "map_data", "NM-1016-3-A50Edge_analyzed_data.csv"))

NM_Line <- read_csv(here("data", "map_data", "NM-1016-3-A10LineScan2_analyzed_data.csv")) %>% 
  mutate(dist=row_number()*0.01)

NM_1016_plot100 <- 
  NM_1016_map100 %>%
  filter(std < 0.15) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(show.legend=FALSE,size=0.5) +
  geom_path(data=NM_Line,aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0, 0.5),
                       breaks = seq(0, 0.4, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
   annotate("rect", 
           xmin = min(NM_1016_map50$x)+15, 
           xmax = max(NM_1016_map50$x)+15, 
           ymin = min(NM_1016_map50$y)-2.3, 
           ymax = max(NM_1016_map50$y)-2, 
           alpha = 0, linetype="longdash",col = "black")+
    labs(x = "mm", y = "mm", fill = "Mg/Ca ")

NM_1016_plot50 <- 
  NM_1016_map50 %>%
 # filter(std < 0.15) %>%
  #filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(show.legend=FALSE,size=0.5) +
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0, 0.4),
                       breaks = seq(0, 0.4, by = 0.1)) +    
theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )+ 
    coord_fixed() +
    labs(x = "mm", y = "mm", colour = "Mg/Ca ")
    
NM_1016_line <- 
  NM_Line %>% 
  filter(relstd<0.1) %>% 
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c() +    
  theme_cowplot(20)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
  
  
NM_1016_plot <- NM_1016_plot100+inset_element(NM_1016_plot50+theme(plot.tag = element_blank()), 0, 0, 0.7, 0.7)
  
  
################
####FIGURE######
################
  (ND_1016_plot+ND_1016_line)/(NM_1016_plot+NM_1016_line)+plot_annotation(tag_levels = "a")
```

Interestingly, the Mg/Ca intensity ratios in the *Nacella* sp. shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g. 0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in ) using the same emission lines. This is particularly visible in @fig-Nac_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in [@Hausmann2023-ih]). While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g. *Patella* shells. Similar observations have been made before [@Graniero2017-io].

Line scans reflect well the changes previously indicated by δ^18^O values and also indicate about two years of recorded growth. That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans. The variability seen in the *P. vulgata* specimens above is not as strong, which might also be related to the fast and consistent growth rate of the shells.

@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their δ^18^O-values (similar graphs for the other *Nacella* sp. specimen can be found in the supplementaries). Compared to the *P. vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and its simple make-up.

```{r Nacella_Compare}
#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
#| fig-height: 10
#| fig-width: 20
#| label: fig-Nac_Comp
 
P1 <-
  ND_Line %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6) %>%
  filter(dist<10) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(20)+
  labs(x = "", y = "Mg/Ca intensity ratio")



P2 <-
  D18O %>% 
  filter(sampleID=="ND-1016-3") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(20)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))


P3 <-
  NM_Line %>% 
  filter(relstd < 0.1) %>%
  filter(dist<9.5) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(20)+
  labs(x = "", y = "Mg/Ca intensity ratio")



P4 <-
  D18O %>% 
  filter(sampleID=="NM-1016-3") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(20)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))


 ((P1 / P2) | (P3 / P4)) + plot_annotation(tag_levels = 'a')

```

The *Nacella* sp. specimens showed similarly clustered correlations, which interestingly were not grouped by species. In fact, specimens ND-1016-3 (*Nacella deaureata)* and NM-1016-3 (*Nacella magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr). Both also share a high coefficient of determination (R^2^) of 0.95 and 0.92 respectively, with the remaining specimens being somewhat lower (R^2^ = 0.76 for ND-1016-4 and R^2^=0.84 for NM-1016-3). The mean R^2^ value for each *Nacella* species is 0.86 for *Nacella deaureata* and 0.88 for *Nacella magellanica*. Why two specimens of different species are so well aligned is not entirely certain and randomness cannot for now be ruled out. That said, since they both have a high R^2^ value, it might just be that they were both minimally affected by factors other than SST, including the sampling location of our line scans.

```{r Nacella_Correlation_Graphs}
#| fig-cap: Correlation graphs for *Nacella* sp. specimens
#| fig-height: 18
#| fig-width: 10
#| label: fig-Nac_Corr


P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(grepl("Nacella", species)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  theme_cowplot(20)+
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
 # ylim(0.05,NA)+
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12"))#+
  # facet_wrap(~region,ncol=2,scales = "free")


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")


P_individual <- P_all+ylim(0,0.4)+
  facet_wrap(~sampleID,ncol=2)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
    formula = y~x
  )+
  theme_cowplot(20)+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )


P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")


```

# Discussion {#Discussion}

The Mg/Ca ratios from the 2D elemental imaging present positive results on SST being the main factor behind Mg/Ca concentration in three Atlantic limpet species. Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data. Nevertheless, the resulting R^2^-values for each species were high (*P. vulgata mean R^2^*=0.89, *N. deaureata mean R^2^*=0.86*, N. magellanica mean R^2^*=0.88). These results are similar to previously published data on *Patella* sp. shells (@tbl-other_corrs). R^2^ values of *P. depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P. caerulea* from the Mediterranean between 0.78 and 0.96 — although with one outlier of 0.33, which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].

*P. rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] @fig-Ferg. Together with the data from two *P. caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used determine a general Mg/Ca-SST relationship with an R^2^-value of 0.79. JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data. While no additional information about those specimens is available, it seems likely that the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well. Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset. Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.

```{r Other_Correlations}
#| tbl-cap: Table showing correlations of Mg/Ca and δ^18^O found in other studies. * indicates that only SST data and no other geochemical data was used. ** these samples were used to determine a shared R^2^-value of 0.79.
#| label: tbl-other_corrs
```

| Species       | Locality       | Specimen | R^2^                    | Study                             |
|-------------|-------------|-------------|-------------|----------------------|
| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
|               |                | LAN545   | 0.86                    |                                   |
|               |                | LAN554   | 0.78                    |                                   |
|               |                | LAN559   | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1   | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2   | 0.84                    |                                   |
|               | Crete          | AF1911A  | 0.91                    |                                   |
|               |                | AF3003A  | 0.92                    |                                   |
|               | Israel         | AKKPC2   | 0.96                    |                                   |
|               |                | AKKPC3   | 0.89                    |                                   |
|               |                | FRMPC1   | 0.84                    |                                   |
|               |                | FRMPC2   | 0.96                    |                                   |
|               | Libya          | MO31A    | 0.83                    |                                   |
|               |                | MP64A    | 0.33                    |                                   |
|               |                | MP67A    | 0.96                    |                                   |
|               |                | MP68A    | 0.81                    |                                   |
|               | Malta          | MA10     | 0.82                    |                                   |
|               | Tunisia        | TUNPC1   | 0.81                    |                                   |
|               |                | TUNPC2   | 0.78                    |                                   |
|               | Turkey         | ANTPC1   | 0.95                    |                                   |
|               |                | ANTPC2   | 0.93                    |                                   |
|               |                | KIZPC1   | 0.94                    |                                   |
|               |                | KIZPC2   | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
|               |                | JL2      | 0.8\*\*                 |                                   |
| *P. caerulea* | Gibraltar      | JM00     | 0.69\*\*                |                                   |
|               |                | JM30     | 0.83\*\*                |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |

```{r Ferguson Data}
#| fig-cap: Correlation graphs for Ferguson et al., 2011 specimens.
#| fig-height: 20
#| fig-width: 10
#| label: fig-Ferg

Ferg_all <-
  Ferguson_data %>% 
  ggplot()+
  aes(d18o,mg_ca,col=shell)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca (mmol/mol)")+
  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))

Ferg_individual <- Ferg_all+
  facet_wrap(~shell,ncol=2)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 1, label.y = 26,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
    formula = y~x
  )+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )

Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
```

Another shell whose previous results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with δ^18^O or SST in implication [@Graniero2017-io]. Different to other specimens of this study, we were able to resample this specimen and getting Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub). These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August). Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88.

```{r ORK_LT5_Subsample}
#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and δ^18^O-values
#| fig-height: 10
#| fig-width: 15
#| label: fig-ORK_sub

# Function to create the desired subsampled tibble
subsample_data <- function(data, window_size = 0.05, step_size = 0.15) {
  max_dist <- max(data$dist)
  subsample_intervals <- seq(0, max_dist, by = step_size)
  
  subsampled_data <- subsample_intervals %>%
    purrr::map_df(~ {
      window_start <- .x
      window_end <- .x + window_size
      
      data_in_window <- data %>%
        filter(dist >= window_start & dist < window_end)
      
      if (nrow(data_in_window) > 0) {
        summarised_data <- data_in_window %>%
          summarise(across(c(x, y, z, mg_ca, std, relstd, dist), mean, .names = "mean_{.col}"))
        summarised_data <- summarised_data %>%
          mutate(window_start = window_start, window_end = window_end)
        return(summarised_data)
      } else {
        return(tibble())
      }
    })
  
  return(subsampled_data)
}

#
# This is data extracted from Fig. 6E of Graniero et al. 2016
GranieroDF <- tibble::tribble(
                ~dist, ~mg_ca,
                  0.1,     19,
                  0.2,   20.6,
                  0.4,   23.8,
                  0.6,   20.4,
                    1,   19.1,
                  1.3,   14.9,
                  1.6,     16,
                  1.9,   17.8,
                  2.3,   15.8,
                  2.8,   17.3,
                  3.1,   13.8,
                  3.3,   24.9,
                  3.7,   16.1,
                    4,   15.6,
                  4.3,   18.4,
                  4.6,   15.5,
                  5.2,   18.6,
                  5.2,   21.2,
                  5.5,   24.1,
                  5.8,   18.7,
                  6.2,   19.3,
                  6.3,   15.5,
                  6.7,   16.7,
                  6.9,   15.3,
                  7.3,   17.9,
                  7.6,   13.7,
                    8,   11.8,
                  8.2,   16.2,
                  8.5,   18.1,
                  8.9,     18,
                  9.1,   19.7,
                  9.4,   20.8,
                  9.7,   19.6,
                   10,   22.2,
                 10.2,   20.2,
                 10.5,   19.3,
                 10.8,   22.8,
                 11.2,   18.7,
                 11.8,   22.1
                )


# Apply the function to your data
MgCa_filtered <- MgCa %>%
  filter(sampleID == "ORK_LT5")

MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)

MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)

P3.1 <-  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca)+
  geom_path(show.legend = FALSE)+
    scale_x_continuous(breaks = seq(0,12,by=2),
                     labels = seq(0,12,by=2))+
  theme_cowplot(20)+
    annotate("rect", xmin = 1.5, xmax = 4, ymin = 0.3, ymax = 0.55, alpha = 0.2, fill = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.3, ymax = 0.55, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
  

# P3.2 <-
#   MgCa_subsampled_300 %>% 
# ggplot()+
#   aes(mean_dist,mean_mg_ca)+
#   geom_path(col="firebrick4")

P3.3 <-
  GranieroDF %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="black")+
   theme_cowplot(20)+
  scale_x_continuous(breaks = seq(0,12,by=2),
                     labels = seq(0,12,by=2))+
  annotate("rect", xmin = 1.5, xmax = 3, ymin = 10, ymax = 25, alpha = 0.2, fill = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.08, ymin = 10, ymax = 25, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
  labs(x = "Distance to edge [mm]", y = "Mg/Ca (mmol/mol)")
  

P3.1/P3.3

```

While we were not able to resample the *Nacella* sp. specimens in Graniero *et al.* [-@Graniero2017-io], we were able to source other modern and archaeological specimens of the same species from Tierra del Fuego, which can serve as an indicator. Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^-values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries). Why these results are different from the studied *Nacella* sp. specimens in Graniero et al. [-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and δ^18^O records of their shells (Figure 6 in [@Graniero2017-io] ) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — fell apart and lead to an overall result of an uncorrelated relationship between Mg/Ca and SST.

# Conclusion

This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relation to sea surface temperature. The results for *Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* were consistently promising with mean R^2^-values of 0.89, 0.86, and 0.88 respectively.

LIBS elemental imaging provided some insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the δ^18^O of the same shells. Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.

These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past. Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a large part of marine mollusc remains in archaeological sites of those regions. Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted δ^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, allows researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.

# References

```{r Modern iso ranges, eval=FALSE}
# Ranges <- D18O %>%
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>%
#   mutate(max=max(d18O),
#          min=min(d18O)) %>%
#   select(min, sampleID,max) %>% unique()
# 
# Modern_Ranges <- Ranges %>%
#   filter(period=="Modern") %>%
#   arrange(range)


# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp. shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.

```

```{r Correlation Graphs, eval=FALSE }
#| fig-cap: Correlation graphs for all specimens
#| fig-height: 20
#| fig-width: 10


P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  mutate(region = substr(sampleID, 1, 1)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
  # facet_wrap(~region,ncol=2,scales = "free")


# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")


P_individual <- P_all+ylim(0,0.8)+
  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
    formula = y~x
  )+
  theme(
    strip.background = element_rect(fill = "white", colour = NA),
    strip.text = element_text(colour = "black"),
    panel.border = element_rect(colour="black")
  )


P_all/P_individual+plot_layout(heights = c(0.5,0.5))


```
